---
title: "EDA"
author: "Ryan Friedman"
format: html
editor: visual
---

## 1. Introduction 

This document will include exploratory data analysis of the water potability dataset.  The goal of the EDA is to understand the structure and quality of the data, as well as any relationships between the variables.  The ultimate goal is to select a model that predicts water potability from the other variables.

```{r}
#| message: false
#| warning: false
library(tidyverse)
library(knitr)
library(gt)
library(ggcorrplot)
```

## 2. Description of Dataset

The dataset for this analysis contains water quality metrics for 3276 different water bodies.

There are nine (9) numerical explanatory variables and one (1) categorical response variable (Potability). The variables and their descriptions are as follows:

| Variable | Type | Description | Units |
|------------------|------------------|-------------------|------------------|
| ph | Numerical | pH of water, 0 to 14 scale | unitless |
| Hardness | Numerical | Amount of dissolved calcium and magnesium | mg/L |
| Solids | Numerical | Total dissolved solids | ppm |
| Chloramines | Numerical | Chloramine concentration | ppm |
| Sulfate | Numerical | Dissolved sulfates | mg/L |
| Conductivity | Numerical | Electrical conductivity | μS/cm |
| Organic_carbon | Numerical | Total organic carbon | ppm |
| Trihalomethanes | Numerical | Trihalomethane concentration | μg/L |
| Turbidity | Numerical | Suspended solids, measured by light-emission | NTU |
| Potability | Categorical | Safe for consumption, 1 = potable, 0 = not potable | unitless |

Further details about the data are available here: <https://www.kaggle.com/datasets/adityakadiwal/water-potability>

For our analysis, Potability is the response variable while the others are designated as predictors.

## 3. Data Cleaning

- Read in the data

```{r}
water_data <- read_csv("Data/water_potability.csv")
```
- Check for any NA values

```{r}
enframe(colSums(is.na(water_data)), # tally the NA values
        name = "Variable", # rename column
        value = "Count") |> # rename column
  gt() |>
  tab_header(title = "NA Values in Original Dataset")
```

It looks like we have quite a few NA values.  Let's look at how many observations are complete.

```{r}
sum(complete.cases(water_data))
```
Before removing the incomplete observations, explore whether the rate of potability among complete observations is similar to the rate among incomplete observations

```{r}
water_data |>
  mutate(status = if_else(complete.cases(water_data), "complete", "incomplete")) |> # assign complete/incomplete status
  group_by(status) |> # group by incomplete/complete
  summarise(count = n(), potability_rate = mean(Potability)) |> # for each group, report count and mean Potability (still a 0 or 1 numeric)
  gt() |>
  tab_header(title = "Summary of Complete vs Incomplete Observations")
```

Potability rates are similar among complete and incomplete observations (40.3% vs 36.9%), indicating that the missing values are not strongly associated with the response.  Therefore we will treat the missing values as approximately random and remove these observations. The remaining dataset of complete observations is sufficiently large (n = 2011) to fit models.

```{r}
water_data <- drop_na(water_data) # removing all observations with NA
```

- Convert Potability to Factor

```{r}
water_data <- water_data |>
  mutate(Potability = factor(Potability))
```

## 4. Exploratory Data Analysis

- First an overview of the center and spread of the predictor variables:

```{r}
summary(water_data)
```

- Now let's examine the predictors grouped by potability

```{r}
#| warning: false
water_data |>
  group_by(Potability) |> # group by potability
  summarise(across(everything(), list(mean = mean, sd = sd), .names = "{.col}&{.fn}")) |> # report mean and std for each variable.  separate names with "&" so as to not conflict with underscore
  pivot_longer(-Potability, # exclude Potability from the pivot
               names_to = c("Variable", ".value"),
               names_sep = "&") |>
  pivot_wider(names_from = Potability,
              values_from = c(mean, sd),
              names_glue = "{.value}_P{Potability}") |> # P0 for non-potable and P1 for potable
  mutate(across(-Variable, round, digits = 2)) |> # rounding to two decimals
  gt() |>
  tab_header(title = "Predictor Means and Standard Deviations by Potability",
             subtitle = "P0 = Non-Potable; P1 = Potable")

```

For each predictor, the mean values across potability classes are very similar.  Standard deviations by potability are comparable as well.  This suggests that a simple linear model is unlikely to be successful in predicting potability.  Tree-based models capture interactions and non-linear effects, and therefore are better suited to detect any relationships that may exist. 

- Let's plot all predictors grouped by potability

```{r}
water_data |>
  pivot_longer(-Potability) |> # required for faceting
  ggplot(aes(value, fill = Potability)) + # coloring by Potability
  geom_density(alpha = 0.4) +
  facet_wrap(~ name, scales = "free") + # facets by name (per the pivot).  "free" allows the scales to adjust for each facet
  labs(title = "Density Plot of Predictors",
       subtitle = "Grouped by Potability",
       fill = "Potability") +
  theme(plot.title = element_text(hjust = 0.5),
        plot.subtitle = element_text(hjust = 0.5)) # Centers the title and subtitle

```

This plot reinforces the earlier finding: the predictor distributions are not clearly separated by potability. Hardness, ph, and Sulfate show slightly different peak heights between the two classes, but their centers and spreads are nearly identical.

- Finally, let's look for any linear correlation between the predictors  

```{r}
water_data |>
  select(-Potability) |> # looking at the predictors without respect to Potability
  cor() |>
  ggcorrplot(type = "lower", lab = TRUE) + 
  labs(title = "Correlation Matrix Heatmap of Predictors") +
  theme(plot.title = element_text(hjust = 0.5))
```

The correlation heatmap reveals negligible linear relationships between any of the predictors. (Note: This does not necessarily mean the predictors are independent, only that there is no linear relationship.)

Next, we will create models to predict potability. 

[Click here for the Modeling Page](Modeling.html)
